Inferring stellar rotation periods using Gaussian processes

### 2016-12-07 documents/ms.tex

\newcommand{\columbia}{1}
\newcommand{\oxford}{2}
\newcommand{\uw}{3}
\newcommand{\princeton}{4}
\newcommand{\naigrain}{333}
\newcommand{\nmcquillan}{100}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}
\newcommand{\acfRMS}{1.45}
\newcommand{\pgramRMS}{1.10}
\newcommand{\mcmcRMS}{0.87}

@@

\begin{abstract}
The light curves of spotted, rotating stars are often non-sinusoidal and
quasi-periodic---spots are not static on the stellar surface and have finite
lifetimes causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid is therefore not a representative model for the
light curve of star with a rotational signal.
Ideally, a physical model of the stellar surface would be conditioned on the
data, however the parameters of such models can be highly degenerate
\citep[\eg][]{Russell1906, Jeffers2009}.
Instead, we use an appropriate {\it effective} model: a Gaussian Process (GP)
with a quasi-periodic covariance kernel function.
By modelling the light curve with a GP, a highly flexible semi-parametric
function, we avoid the necessity of choosing a `best fitting' functional form,
whilst sampling directly from the posterior Probability Density Function (PDF)
of the periodic parameter and marginalising over the other kernel
hyperparameters.
We used \naigrain\ simulated light curves with a range of rotation periods
to test the GP model.
We attempted to recover the rotation periods using three methods: our GP
method, a sine-fitting periodogram method and an AutoCorrelation Function
(ACF) method.
The posterior PDF of the rotation period parameter was sampled using the
affine invariant ensemble Markov Chain Monte Carlo (MCMC) sampler {\tt emcee}
\citep{Foreman-Mackey2013}, and the GP operations were performed using the
{\tt george} python package \citep{George}.
Rotation periods inferred via this method are more precise and accurate than
the periodogram and ACF methods.
Furthermore, rotation periods are inferred probabilistically, allowing them to
be easily incorporated into follow-on studies, inferring stellar ages or
characterising star-planet interactions for example.
In addition, unlike the sine-fitting periodogram and ACF methods, the
posterior PDF samples allow for proper characterisation of uncertainties.
This method eliminates the need to define a heuristic detection criterion such
as are often used in the literature: a light curve that does not contain a
rotation signal will return the prior PDF.

@@

The light curves of spotted, rotating stars are often non-sinusoidal and
Quasi-Periodic (QP).
These stars vary in brightness due to active regions on their surfaces which
rotate in and out of view.
The non-sinusoidal quality is caused by the complicated surface spot patterns
and the quasi-periodicity is produced both by the finite lifetimes of these
active regions and the presence of differential rotation on the stellar
surface.
A strictly periodic sinusoid is therefore not a good model for stellar light
curves.
In an ideal world, a physical model of the stellar surface would be
conditioned on the data.
A physically realistic model would perfectly capture the complexity of shapes
within stellar light curves as well as the quasi-periodic nature, allowing for
extremely precise probabilistic period recovery.
However, such physical models require many free parameters in order to
accurately represent a stellar surface and some of these parameters are
extremely degenerate \citep[\eg][]{Russell1906, Jeffers2009}.
In addition to global stellar parameters such as inclination and rotation
period, each spot or active region should have (at minimum) a longitude,
latitude, size, temperature and lifetime.
Considering that many stars have on the order of hundreds of spots, the number
of free parameters quickly becomes unwieldy, especially if the posterior
Probability Density Functions (PDFs) of these parameters are explored with
MCMC.
Simplified spot models, such as the one described in \citet{Lanza2014} where
only two spots are modelled, have produced successful results, however these
simplified models sacrifice some precision due to lack of model flexibility.
Instead of using a physical model for stellar light curves, we choose to use
an {\it effective} model: one which captures the behaviour but is not
physically motivated, although the parameters of this model may be {\it
interpreted} as physical ones.
An ideal effective model for the light curve of a spotted, rotating star is
one with a small number of non-degenerate parameters that is flexible enough
to perfectly capture non-sinusoidal, QP behaviour.
These requirements are fulfilled by a Gaussian process (GP) model.

The standard methods used for measuring rotation periods include detecting
peaks in a Lomb-Scargle \citep{Lomb1976, Scargle1982} (LS) periodogram
\citep[e.g.][]{Reinhold2013}, Auto-Correlation Functions (ACFs)
\citep{Mcquillan2013} and wavelet transforms \citep{Garcia2014}.
The precision of the LS periodogram and wavelet methods are limited by the
suitability of the model choice.
A sinusoid is used in the case of the LS periodogram and the wavelet method
relies on a choice of mother wavelet that is assumed to describe the data over
a range of transpositions \citep[see, \eg][]{Carter2010}.
In contrast, the ACF method is much better suited to signals that are
non-sinusoidal.
In fact, it doesn't matter what shape the signal is: as long as it is
approximately periodic the ACF will display peaks located at the rotation
period.
A drawback of the ACF method however, is that it requires data to be
evenly-spaced\footnote{\citet{Edelson1988} describe a method for computing
ACFs for unevenly-spaced data.} which is not the case with \Kepler\ light
curves (although in many cases it can be approximated as uniformly sampled).
An ACF is also an operation performed on the data, not a generative model of
the data and is therefore not inherently probabilistic.
For this reason it is very difficult to estimate the uncertainty on an ACF
rotation period measurement.
Many rotation periods in the literature have been inferred by measuring the
position of the first peak in an ACF, however this approach can be dangerous.
The exponential decay in correlation can shift the peak position short-wards
of its true value, leading to an underestimate of the period.
We return to this point in section \textsection \ref{sec:discussion}.
% In order to avoid this, a period should be inferred by modelling the entire
% ACF, not just measuring the position of the first peak.

The motivation for developing a GP-based method for rotation period inference
is, firstly to measure more accurate and precise rotation periods using a
better-suited generative model than a sinusoid for the reasons explained
above.
Secondly, in order to infer {\it probabilistic} periods, i.e. to estimate the
posterior PDF of the rotation period and thereby measure a realistic and
representative uncertainty.
Thirdly, to allow for an additional correlated noise model to be included
during regression, the parameters of which could be marginalised over.
% And fourthly to provide some way of determining whether a periodic model is
% supported by the data over a purely stochastic one.

@@

GP models parameterise the covariance between data points and a kernel
function provides the form of covariance matrix parameterisation.
For example, take the time-series in figure \ref{fig:GP_example}.
This is the \kepler\ light curve of KIC 1430163, a relatively active star that
rotates once every $\sim$ 4 days.
The stochastic variability in this time-series is typical of \kepler\ FGK
stars.
Clearly, data points in this light curve are correlated.
Points that are close together in time are tightly correlated and those that
are widely separated in time are loosely correlated.
The way in which the covariance varies with the separation between data points
is modelled when using GP regression; it is not the data but the {\it
covariance structure} that is modelled.
This fact is what gives GPs their flexibility---they can model any time
series with a similar covariance structure.
In addition, a very simple function can usually capture the covariance
structure of a light curve, whereas a much more complicated one may be
required to model the time-series itself.
% The light curve of \Kepler-452 has been modelled with a GP in figure
% \ref{fig:GP_example}, shown in orange.
In figure \ref{fig:GP_example} the light curve of KIC 1430163 has been
modelled with a GP.
% Both provide adequate fits to the data, however only the periodic kernel
% function, `QP' is a useful model because it has a periodic parameter.
% I return to this point shortly.

A range of kernel functions could be used to describe stellar variability.
For example, the simplest and most commonly used kernel function, the `Squared
Exponential' (SE) produces an adequate fit to the KIC 1430163 light curve.
The SE kernel function is defined as,
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} \right).
\end{equation}
\label{eq:SE}
Here $A$ is the amplitude of covariance, $l$ is the length scale of
exponential decay and $x_i-x_j$ is the separation between data points.
The SE kernel function has the advantage of being very simple---it has just
two parameters, a covariance amplitude and length scale: $A$ and $l$.
If $l$ is large two data points far apart in $x$ will be tightly correlated,
and if small they will be loosely correlated.
Another property of the SE kernel function is that it produces functions that
are infinitely differentiable.
It is therefore possible to model a data set and its derivatives
simultaneously.
% \citep[see \eg][]{Rajpaul2015}
The SE kernel function may be an adequate model of the covariance in stellar
light curves, but it is not a {\it useful} one for the problem of rotation
period inference because it does not have a parameter that describes a period.
In order to infer rotation periods it is necessary to use a periodic kernel
function.
For this reason, we use the `Quasi-Periodic' kernel function.
The QP kernel function is defined as
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} -
\Gamma^2 \sin^2(\frac{2\pi}{P}) \right).
\end{equation}
\label{eq:QP}
It is the product of the SE kernel function, which describes the overall
covariance decay, and an exponentiated, squared, sinusoidal kernel function
that describes the periodic covariance structure.
$P$ can be interpreted as the rotation period of the star and $\Gamma$
controls the amplitude of the sinusoidal term.
This kernel function allows two data points that are separated in time by one
rotation period to be tightly correlated, while points separated by half a
period can be weakly correlated.
Both the QP kernel function was used to produce the model shown in figure
~\ref{fig:GP_example}.

In order to infer a stellar rotation period from a light curve, we fit a GP
model with a QP kernel function to the data.
The likelihood of the model, conditioned on the data could then be maximised
in order to find the maximum likelihood value for $P$.
In this study however, the full posterior PDFs of the parameters are explored
using MCMC.
This approach comes at a cost: a GP model is expensive to compute once as it
requires a matrix inversion and determinant evaluation, let alone the many
thousands of times that is necessary to fully explore the posteriors of the
parameters.
However, fully exploring the posterior PDF of $P$ allows for an accurate
estimate of the uncertainty on the rotation period.

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/001430163.pdf}
\caption[A light curve with a GP model.]
{Light curve of KIC 1430163, an active star with a rotation period of $\sim$ 4
days.
The blue line shows a fit to the data using a Gaussian process model with a QP
covariance kernel function.}
\label{fig:GP_example}
\end{center}
\end{figure}

\section{Rotation period injection and recovery}

In order to test the GP method we attempted to the measure rotation periods of
a set of simulated light curves.
We used \naigrain\ light curves generated for the \citet{Aigrain2015}
`hare and hounds' rotation period recovery experiment.
These light curves were simulated by placing dark, circular spots with slowly
evolving size, on the surface of bright, rotating spheres, ignoring
limb-darkening effects.
\citet{Aigrain2015} simulated one thousand light curves in order to test the
ability of participating teams to recover both the stellar rotation periods
and the rotational shear: the amplitude of surface differential rotation.
% Since we are not interested in recovering differential rotation we only used
% those light curves simulated {\it without} differential rotation, of which
% there are \naigraincurves.
For the purpose of demonstrating that our method can recover rotation periods,
we selected only the \naigrain\ out of the 1000 without differential
rotation.
We opted to use only solid-body rotators because differential rotation may
produce some additional scatter in the rotation period measurements recovered.
In future we intend to test whether we can recover differential rotation using
the GP method and will then use the full set of 1000 light curves.
The distributions of input parameters used to generate these light curves are
listed in \citet{Aigrain2015}.
Light curves were simulated with a real \Kepler\ long-cadence time array:
one data point every thirty minutes over a four year duration.
90\% of the rotation periods of the simulations were randomly drawn from a
log-uniform distribution between 10 and 50 days and 10\% from a log-uniform
distribution between 1 and 10 days.
The light curves were also generated with a range of stellar inclination
angles, activity levels, spot lifetimes and more.
Once simulated, the light curves were added to real \kepler\ light curves with
no obvious astrophysical variability in order to preserve the noise properties
of the time series.
Figure \ref{fig:demo_lc} shows an example of a simulated, noise-free
light curve with a period of 17.04 days.

@@

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/period_hist.pdf}
\caption{A histogram of the rotation periods used to generate the simulated
light curves in \citet{Aigrain2015}.}
\label{fig:period_hist}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/demo_lc.pdf}
\caption[A simulated light curve.]
{An example simulated light curve. This `star' has a rotation period of 20.8
days.}
\label{fig:demo_lc}
\end{center}
\end{figure}

We attempted to recover the rotation periods of these \naigrain\ light
curves using three rotation period recovery methods: the ACF method, the LS
periodogram method and the GP method.

\subsection{ACF}

We calculated an ACF for each light curve following the method of
\citet{Mcquillan2013}.
In this method, an ACF is calculated for each light curve and smoothed by
convolving with a Gaussian with $\sigma=9$ days.
A rotation period is estimated as the lag-time of the highest peak in the ACF,
less than 100 days.
This is not always the first peak: the second can be larger than the first if
there are two active regions at or near opposite longitudes on the surface of
the star---this would produce a light curve with two dips per rotation period.
We restrict our search to periods less than 100 days because this is the
maximum period that is realistically recoverable in \kepler\ light curves ---
instrumental noise significantly distorts the signal on longer timescales.
An example ACF of the light curve in figure~\ref{fig:demo_lc} is shown
in figure \ref{fig:demo_acf}.

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/demo_ACF.pdf}
\caption[ACF of a simulated light curve.]
{An autocorrelation function of the simulated light curve shown in figure
\ref{fig:demo_lc}.
The vertical blue line shows the period measured using the ACF method (20.4
days) and the pink dashed line shows the period that was used to simulate the
light curve (20.8 days).}
\label{fig:demo_acf}
\end{center}
\end{figure}

@@

We applied the ACF method to the sample of \naigrain\ simulated light
curves.
Periods measured using the ACF method are plotted against the original
rotation period values used to generate the light curves in figure
\ref{fig:compare_acf}.
The $2\times$ and $1/2\times$ harmonic lines are shown as dashed lines.
The RMS scatter about the $x=y$ line for rotation periods measured using the
ACF method is \acfRMS.
The agreement between the injected and recovered rotation periods is good in
general, however twice the true rotation period is often found.
In addition, some rotation periods are drastically over- and underestimated.
Of the points clustered close to the $x=y$ line, more slightly below it than
above it, \ie\ rotation periods are preferentially underestimated.
This is a feature of the peak position measurement method that is performed on
the ACF.
ACFs of stellar light curves are, like the QP kernel function in equation
\ref{eq:QP}, periodic functions added to decaying exponentials.
In such functions the peak positions can be shifted towards the left, the
short period end, because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect, however the standard practice is to
simply measure the peak position without taking it into account.
We adopt the standard practice approach here in order to provide a comparison
between our new method and the current standard procedure.

@@

\subsection{The sine-fitting periodogram method}

For each simulated light curve, a Lomb-Scargle (LS) periodogram
\citep{Lomb1976, Scargle1982} was computed over a grid of 10,000 periods,
evenly spaced in frequency, between 1 and 100 days\footnote{LS periodograms
were calculated using the gatspy Python module:
\url{https://github.com/astroML/gatspy/tree/master/gatspy/periodic}.}
The period of the highest peak in the periodogram was adopted as the rotation
period.
The resulting recovered rotation periods are plotted as a function of the true
periods in figure \ref{fig:pgram_compare}.
These recovered rotation periods are in general more accurate than the ACF
results: they do not systematically over or under predict rotation period.
The RMS scatter for rotation periods measured using the LS periodogram
technique is \pgramRMS.
The longer-period measurements provide the dominant contribution to this RMS.
% We believe that the unsuccessful recoveries occurring preferentially for
% shorter rotation periods are produced by additional, longer timescale
% variations in some of the light curves.
% Specifically, by fluctuations produced by changes in the total spot coverage
% on the stellar surface.
% These fluctuations can occasionally be larger in magnitude than those produced
% by the rotational signal itself and are present in the majority of outlying
% light curves that occupy the top left of figure
% \ref{fig:pgram_compare_noise-free}.

@@

$\sigma$ is an additional white noise term added to the diagonal elements of
the covariance matrix. It is the fraction by which the observational
uncertainties have been underestimated.
If the uncertainties reported on the data are too small, this parameter will
be non-zero.
In practice however, this parameter should always be non-zero when performing
GP regression.
This is because the covariance matrix must be positive definite, however
matrix inversion performed using most solvers is approximate, not exact,
therefore slight deviations from positive definiteness can arise.
Including a small amount of extra variance in the model allows enough
flexibility that the matrix inversion algorithms do not run into numerical
issues.

@@

\subsection{Real \kepler\ data}

>>>>

\newcommand{\columbia}{1}
\newcommand{\oxford}{2}
\newcommand{\uw}{3}
\newcommand{\princeton}{4}
\newcommand{\naigrain}{333}
\newcommand{\nmcquillan}{100}
\newcommand{\kepexample}{1430163}
\newcommand{\kepexampleperiod}{4}
\newcommand{\aigrainexampleperiod}{20.8}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}
\newcommand{\acfRMS}{1.45}
\newcommand{\pgramRMS}{1.10}
\newcommand{\mcmcRMS}{0.87}

@@

\begin{abstract}
The light curves of spotted, rotating stars are often non-sinusoidal and
quasi-periodic---spots are not static on the stellar surface and have finite
lifetimes causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid is therefore not a representative model for the
light curve of star with a rotational signal.
Ideally, a physical model of the stellar surface would be conditioned on the
data, however the parameters of such models can be highly degenerate
\citep[\eg][]{Russell1906, Jeffers2009}.
Instead, we use an appropriate {\it effective} model: a Gaussian Process (GP)
with a quasi-periodic covariance kernel function.
By modelling the light curve with a GP, a highly flexible semi-parametric
function, we avoid the necessity of choosing a `best fitting' functional form,
whilst sampling directly from the posterior Probability Density Function (PDF)
of the periodic parameter and marginalising over the other kernel
hyperparameters.
We used \naigrain\ simulated light curves with a range of rotation periods
to test the GP model.
We attempted to recover the rotation periods using three methods: our GP
method, a sine-fitting periodogram method and an AutoCorrelation Function
(ACF) method.
The posterior PDF of the rotation period parameter was sampled using the
affine invariant ensemble Markov Chain Monte Carlo (MCMC) sampler {\tt emcee}
\citep{Foreman-Mackey2013}, and the GP operations were performed using the
{\tt george} python package \citep{George}.
Rotation periods inferred via this method are more precise and accurate than
the periodogram and ACF methods.
Furthermore, rotation periods are inferred probabilistically, allowing them to
be easily incorporated into follow-on studies; inferring stellar ages or
characterising star-planet interactions for example.
In addition, unlike the sine-fitting periodogram and ACF methods, the
posterior PDF samples allow for proper characterisation of uncertainties.
This method eliminates the need to define a heuristic detection criterion such
as are often used in the literature: a light curve that does not contain a
rotation signal will return the prior PDF.

@@

The light curves of spotted, rotating stars are often non-sinusoidal and
Quasi-Periodic (QP).
These stars vary in brightness due to active regions on their surfaces which
rotate in and out of view.
The non-sinusoidal quality is caused by the complicated surface spot patterns
and the quasi-periodicity is produced both by the finite lifetimes of these
active regions and the presence of differential rotation on the stellar
surface.
A strictly periodic sinusoid is therefore not a good model for stellar light
curves.
In an ideal world, a physical model of the stellar surface would be
conditioned on the data.
A physically realistic model would perfectly capture the complexity of shapes
within stellar light curves as well as the quasi-periodic nature, allowing for
extremely precise probabilistic period recovery.
However, such physical models require many free parameters in order to
accurately represent a stellar surface and some of these parameters are
extremely degenerate \citep[\eg][]{Russell1906, Jeffers2009}.
In addition to global stellar parameters such as inclination and rotation
period, each spot or active region should have (at minimum) a longitude,
latitude, size, temperature and lifetime.
Considering that many stars have on the order of hundreds of spots, the number
of free parameters quickly becomes unwieldy, especially if the posterior PDFs
of these parameters are explored with MCMC.
Simplified spot models, such as the one described in \citet{Lanza2014} where
only two spots are modelled, have produced successful results, however these
simplified models sacrifice some precision due to lack of model flexibility.
Instead of using a physical model for stellar light curves, we choose to use
an {\it effective} model: one which captures the behaviour but is not
physically motivated, although the parameters of this model may be {\it
interpreted} as physical ones.
An ideal effective model for the light curve of a spotted, rotating star is
one with a small number of non-degenerate parameters that is flexible enough
to perfectly capture non-sinusoidal, QP behaviour.
These requirements are fulfilled by a Gaussian process (GP) model.

The standard methods used for measuring rotation periods include detecting
peaks in a Lomb-Scargle \citep{Lomb1976, Scargle1982} (LS) periodogram
\citep[e.g.][]{Reinhold2013}, Auto-Correlation Functions (ACFs)
\citep{Mcquillan2013} and wavelet transforms \citep{Garcia2014}.
The precision of the LS periodogram and wavelet methods are limited by the
suitability of the model choice.
A sinusoid is used in the case of the LS periodogram and the wavelet method
relies on a choice of mother wavelet that is assumed to describe the data over
a range of transpositions \citep[see, \eg][]{Carter2010}.
In contrast, the ACF method is much better suited to signals that are
non-sinusoidal.
In fact, it doesn't matter what shape the signal is: as long as it is
approximately periodic the ACF will display peaks located at the rotation
period.
A drawback of the ACF method however, is that it requires data to be
evenly-spaced\footnote{\citet{Edelson1988} describe a method for computing
ACFs for unevenly-spaced data.} which is not the case with \Kepler\ light
curves (although in many cases it can be approximated as uniformly sampled).
An ACF is also an operation performed on the data, not a generative model of
the data and is therefore not inherently probabilistic.
For this reason it is very difficult to estimate the uncertainty on an ACF
rotation period measurement.
% Many rotation periods in the literature have been inferred by measuring the
% position of the first peak in an ACF, however this approach can be dangerous.
% The exponential decay in correlation can shift the peak position short-wards
% of its true value, leading to an underestimate of the period.
% We return to this point in section \textsection \ref{sec:discussion}.

The motivation for developing a GP-based method for rotation period inference
is, firstly to measure more accurate and precise rotation periods using a
better-suited generative model than a sinusoid for the reasons explained
above.
Secondly, in order to infer {\it probabilistic} periods, i.e. to estimate the
posterior PDF of the rotation period and thereby produce a realistic and
estimate of its uncertainty.
Thirdly, to allow for an additional correlated noise model to be included
during regression, the parameters of which could be marginalised over.
% And fourthly to provide some way of determining whether a periodic model is
% supported by the data over a purely stochastic one.

@@

GP models parameterise the covariance between data points and a kernel
function provides the form of covariance matrix parameterisation.
For example, take the time-series in figure \ref{fig:GP_example}.
This is the \kepler\ light curve of KIC \kepexample, a relatively active star that
rotates once every $\sim$ 4 days.
The stochastic variability in this time-series is typical of \kepler\ FGK
stars.
Clearly, data points in this light curve are correlated.
Points that are close together in time are tightly correlated and those that
are widely separated in time are loosely correlated.
The way in which the covariance varies with the separation between data points
is modelled when using GP regression; it is not the data but the {\it
covariance structure} that is modelled.
This fact is what gives GPs their flexibility---they can model any time
series with a similar covariance structure.
In addition, a very simple function can usually capture the covariance
structure of a light curve, whereas a much more complicated one may be
required to model the time-series itself.
% The light curve of \Kepler-452 has been modelled with a GP in figure
% \ref{fig:GP_example}, shown in orange.
In figure \ref{fig:GP_example} the light curve of KIC \kepexample\ has been
modelled with a GP.
% Both provide adequate fits to the data, however only the periodic kernel
% function, `QP' is a useful model because it has a periodic parameter.
% I return to this point shortly.

A range of kernel functions could be used to describe stellar variability.
For example, the simplest and most commonly used kernel function, the `Squared
Exponential' (SE) produces an adequate fit to the KIC \kepexample\ light
curve.
The SE kernel function is defined as,
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} \right).
\end{equation}
\label{eq:SE}
Here $A$ is the amplitude of covariance, $l$ is the length scale of
exponential decay and $x_i-x_j$ is the separation between data points.
The SE kernel function has the advantage of being very simple---it has just
two parameters, a covariance amplitude and length scale: $A$ and $l$.
If $l$ is large two data points far apart in $x$ will be tightly correlated,
and if small they will be loosely correlated.
Another property of the SE kernel function is that it produces functions that
are infinitely differentiable.
It is therefore possible to model a data set and its derivatives
simultaneously.
% \citep[see \eg][]{Rajpaul2015}
The SE kernel function is not a good model of the covariance in stellar light
curves, nor a {\it useful} one for the problem of rotation period inference
because it does not capture periodic behaviour.
In order to infer rotation periods it is necessary to use a periodic kernel
function.
For this reason, we use the `Quasi-Periodic' kernel function.
The QP kernel function is defined as
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} -
\Gamma^2 \sin^2(\frac{2\pi}{P}) \right).
\end{equation}
\label{eq:QP}
It is the product of the SE kernel function, which describes the overall
covariance decay, and an exponentiated, squared, sinusoidal kernel function
that describes the periodic covariance structure.
$P$ can be interpreted as the rotation period of the star and $\Gamma$
controls the amplitude of the $\sin^2$ term.
If $\Gamma$ is very small, only points almost exactly one period away are
tightly correlated and points that are slightly more or less than one period
away are very loosely correlated.
If $\Gamma$ is large, points separated by one period are tightly
correlated and points separated by slightly more or less are still highly
correlated although less so.
This kernel function allows two data points that are separated in time by one
rotation period to be tightly correlated, while points separated by half a
period can be weakly correlated.
We also use an extra parameter, $\sigma$ which is an additional white noise
term added to the diagonal elements of the covariance matrix.
It is the fraction by which the observational uncertainties have been
underestimated --- if the uncertainties reported on the data are too small,
this parameter will be non-zero.
In practice however, this parameter should always be non-zero when performing
GP regression.
This is because the covariance matrix must be positive definite, however
matrix inversion performed using most solvers is approximate, not exact,
therefore slight deviations from positive definiteness can arise.
Including a small amount of extra variance in the model allows enough
flexibility that the matrix inversion algorithms do not run into numerical
issues.

This QP kernel function was used to produce the model shown in figure
~\ref{fig:GP_example}.

In order to infer a stellar rotation period from a light curve, we fit a GP
model with a QP kernel function to the data.
The likelihood of the model, conditioned on the data could then be maximised
in order to find the maximum likelihood value for $P$.
In this study however, the full posterior PDFs of the parameters are explored
using MCMC.
This approach comes at a cost: a GP model is expensive to compute once as it
requires an inversion and determinant evaluation of the covariance matrix, let
alone the many thousands of times that is necessary to fully explore the
posteriors of the parameters.
Fully exploring the posterior PDF of $P$ is important however, as it provides
an accurate estimate of the uncertainty on the rotation period.

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/001430163.pdf}
\caption[A light curve with a GP model.]
{Light curve of KIC \kepexample, an active star with a rotation period of
$\sim$ \kepexampleperiod\ days.
The blue line shows a fit to the data using a Gaussian process model with a QP
covariance kernel function.}
\label{fig:GP_example}
\end{center}
\end{figure}

In \textsection \ref{sec:simulations} of this paper we describe the simulated
light curves used to test rotation period inference methods.
In \textsection \ref{sec:method} we describe the various methods used to
recover the rotation periods: the periodogram, ACF and GP methods.
In \textsection \ref{sec:kepler} we apply our method to real \kepler\ data and
the results of the paper are discussed in \textsection \ref{sec:discussion}.

\section{Simulated light curves}
\label{sec:simulations}

In order to test the GP method we attempted to the measure rotation periods of
a set of simulated light curves.
We used \naigrain\ light curves generated for the \citet{Aigrain2015}
`hare and hounds' rotation period recovery experiment.
These light curves were simulated by placing dark, circular spots with slowly
evolving size, on the surface of bright, rotating spheres, ignoring
limb-darkening effects.
\citet{Aigrain2015} simulated one thousand light curves in order to test the
ability of participating teams to recover both the stellar rotation periods
and the rotational shear: the amplitude of surface differential rotation.
% Since we are not interested in recovering differential rotation we only used
% those light curves simulated {\it without} differential rotation, of which
% there are \naigraincurves.
For the purpose of demonstrating that our method can recover rotation periods,
we selected only the \naigrain\ out of the 1000 without differential
rotation.
We opted to use only solid-body rotators because differential rotation may
produce some additional scatter in the rotation period measurements recovered.
In future we intend to test whether we can recover differential rotation using
the GP method and will then use the full set of 1000 light curves.
The distributions of input parameters used to generate these light curves are
listed in \citet{Aigrain2015}.
Light curves were simulated with a real \Kepler\ long-cadence time array:
one data point every thirty minutes over a four year duration.
90\% of the rotation periods of the simulations were randomly drawn from a
log-uniform distribution between 10 and 50 days and 10\% from a log-uniform
distribution between 1 and 10 days.
A histogram of the \naigrain\ solid-body rotation periods is shown in figure
\ref{fig:period_hist}.
The light curves were also generated with a range of stellar inclination
angles, activity levels, spot lifetimes and more.
Once simulated, the light curves were added to real \kepler\ light curves with
no obvious astrophysical variability in order to preserve the noise properties
of the time series.
Figure \ref{fig:demo_lc} shows an example of a simulated light curve with a
period of \aigrainexampleperiod days.

@@

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/period_hist.pdf}
\caption{A histogram of the rotation periods used to generate the \naigrain\
simulated light curves in \citet{Aigrain2015}.}
\label{fig:period_hist}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/demo_lc.pdf}
\caption[A simulated light curve.]
{An example simulated light curve. This `star' has a rotation period of
\aigrainexampleperiod\ days.}
\label{fig:demo_lc}
\end{center}
\end{figure}

% We attempted to recover the rotation periods of these \naigrain\ light curves
% using three methods: the ACF method; the LS periodogram method; and the GP
% method, and compared their performances.

\section{Rotation period recovery}
\label{sec:method}
\subsection{ACF}

We calculated an ACF for each light curve following the method of
\citet{Mcquillan2013}.
An ACF is calculated for each light curve and smoothed by convolving with a
Gaussian with $\sigma=9$ days.
A rotation period is estimated as the lag-time of the highest peak in the ACF,
less than 100 days.
This is not always the first peak: the second can be larger than the first if
there are two active regions at or near opposite longitudes on the surface of
the star---producing a light curve with two dips per rotation period.
We restrict our search to periods less than 100 days because this is the
maximum period that is realistically recoverable in \kepler\ light curves ---
instrumental noise significantly distorts signals on longer timescales.
An example ACF of the light curve in figure~\ref{fig:demo_lc} is shown
in figure \ref{fig:demo_acf}.

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/demo_ACF.pdf}
\caption[ACF of a simulated light curve.]
{An autocorrelation function of the simulated light curve shown in figure
\ref{fig:demo_lc}.
The vertical blue line shows the period measured using the ACF method (20.4
days) and the pink dashed line shows the period that was used to simulate the
light curve (\aigrainexampleperiod\ days).}
\label{fig:demo_acf}
\end{center}
\end{figure}

@@

We applied the ACF method to the sample of \naigrain\ simulated light
curves.
Periods measured using the ACF method are plotted against the original
rotation period values used to generate the light curves in figure
\ref{fig:compare_acf}.
The $2n$ and $\frac{1}{2}n$ harmonic lines are shown as dashed lines.
The RMS scatter about the $x=y$ line for rotation periods measured using the
ACF method is \acfRMS.
The agreement between the injected and recovered rotation periods is good in
general, however twice the true rotation period is often found.
In addition, some rotation periods are drastically over- and underestimated.
Of the points clustered close to the $x=y$ line, more fall slightly below it
than above it, \ie\ rotation periods are preferentially underestimated.
This is a feature of the peak position measurement method that is performed on
the ACF.
ACFs of stellar light curves are similar to the QP kernel function in equation
\ref{eq:QP}: periodic functions added to decaying exponentials.
In such functions the peak positions can be shifted towards the left, the
short period end, because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect, however the standard practice is to
simply measure the peak position without taking it into account.
We adopt the standard practice approach here in order to provide a comparison
between our new method and that used largely in the literature.

@@

\subsection{LS periodogram}

For each simulated light curve, a LS periodogram
\footnote{LS periodograms were calculated using the gatspy Python module:
\url{https://github.com/astroML/gatspy/tree/master/gatspy/periodic}.}
was computed over a grid of 10,000 periods, evenly spaced in frequency,
between 1 and 100 days.
The period of the highest peak in the periodogram was adopted as the rotation
period.
The resulting recovered rotation periods are plotted as a function of the true
periods in figure \ref{fig:pgram_compare}.
The RMS scatter for rotation periods measured using the LS periodogram
technique is \pgramRMS.
The longer-period measurements provide the dominant contribution to this RMS.
% We believe that the unsuccessful recoveries occurring preferentially for
% shorter rotation periods are produced by additional, longer timescale
% variations in some of the light curves.
% Specifically, by fluctuations produced by changes in the total spot coverage
% on the stellar surface.
% These fluctuations can occasionally be larger in magnitude than those produced
% by the rotational signal itself and are present in the majority of outlying
% light curves that occupy the top left of figure
% \ref{fig:pgram_compare_noise-free}.

@@

@@

\section{Real \kepler\ data}
\label{sec:kepler}
